Symbolic Trajectories

### 2012-08-17 Algebras/SymbolicTrajectory/st_impl.tex
*** empty log message ***

\subsubsection{Class NFA}
The NFA instance (there is only one) is constructed in several steps. First, the transition function $\delta$ is computed. We model $\delta$ as a matrix of integer sets, i.e., for every pair $(i,j)$ of an NFA state and a unit pattern id, there is a (possibly empty) set of states that become active if a unit label matching the unit pattern $p_j$ is read in state $i$ (replacing the formerly active states). After this, a vector of sets of numbers is filled with all proper matches, e.g., if unit pattern $p_m$ matches unit label $n$, the number $n$ is inserted into the $m$th set.

@@

@@

\subsubsection{Preparing the Condition Evaluation}\label{sec:preparing the condition evaluation}
Since the conditions are evaluated by \secondo, there are only two rules they have to obey:
\begin{itemize}
  \item at least one term containing a variable and a condition type (e.g., \texttt{Y.time}) is mandatory for every condition
  \item every condition must be a boolean expression, i.e., its evaluation has to yield \ic{true} or \ic{false}
\end{itemize}

@@

For realizing this, we need to compute a set of cardinality candidates $C_i$ for every unit pattern id $i$, which is done with the help of the matching sets ($M_s$ in algorithm \ref{alg:match}). We distinguish three cases:
\begin{itemize}
  \item a simple pattern: $C_i=\{1\}$
  \item a sequence pattern surrounded by two proper matches: $C_i=M_{i+1}-M_{i-1}$\footnote{For two sets $A, B\subseteq \mathbb{N}_0$, we denote the set of all non-negative differences $\{b-a\colon a\in A, b\in B, a\leq b\}$ by $B-A$}
  \item a sequence pattern besides other sequence patterns: $C_{i}=\{1,\dots,\max\{M_{n}-M_{p}\}\}$, with $n$ and $p$ denoting the next and the previous position of simple patterns, respectively
\end{itemize}

@@

These candidate sets enable us to build further sets providing an evaluation order, as done in algorithm \ref{alg:buildsequences}. First, all cardinality set sizes except the largest one are multiplied, yielding the number of matching sequence candidates which are going to be built in the next step. We skip the largest set in the multiplication since a matching sequence is already determined by $n-1$ numbers. This technique reduces the computational cost for the outer loop by a factor up to $\left|ml\right|$, depending on the occurrence of $+$ and $\ast$ in the pattern.

\begin{algorithm}[ht]
  \caption{\label{alg:buildsequences}\ic{buildSequences$(C)$}}
    \KwIn{$C$ - a vector of sets containing the cardinality candidates for every unit pattern.}
    \KwOut{$S$ - a set of multisets representing all possible matching sequences.}
    let $m\in \{0,\dots,n-1\}$, with $\left|C_m\right|=\max\limits_i \left|C_i\right|$\;
    let $size=\prod_{0\leq i<n,i\neq m}\left|C_i\right|$\;
    \For{$i=0$ \KwTo $size-1$}{
      let $j=i$\;
      let $sq$ be a multiset of integers, initially empty\;
      let $cards$ be a vector of $n$ integers, initially empty\;
      let $cSum=0$\;
      \ForEach{$s\in\{0,\dots,n-1\}\setminus \{m\}$}{
        $cards_s=C_{s,\,j\ \textnormal{mod}\ \left|C_s\right|}$\;
        $cSum=cSum+C_{s,\,j\ \textnormal{mod}\ \left|C_s\right|}$\;
        $j=j/\left|C_s\right|$\;
        \lIf(\tcp*[f]{\scriptsize{sum too high}}){$cSum>\left|ml\right|$}{$s=n$}
      }
      \If{$\left|ml\right|-cSum\in C_m$}{
        $cards_m=\left|ml\right|-cSum$\;
        $sq=\{0,\sum_{k=0}^{0}cards_k,\dots,\sum_{k=0}^{n-2}cards_k\}$\;
        $S\leftarrow sq$\;
      }
    }
\end{algorithm}

@@

Since algorithm \ref{alg:buildsequences} is heavily time-consuming, this version is only executed if the operator \texttt{rewrite} is called. The other case is discussed in \ref{sec:condition evaluation}.

\subsubsection{Condition Evaluation}\label{sec:condition evaluation}
The next step for \texttt{rewrite} is to scan $S$, eliminating the sequences which do not fulfill the set of specified conditions in order to obtain the final set of sequences for rewriting the pattern.

However, we suppose the operator \texttt{matches} is invoked, with the consequence that it suffices to verify whether there is a sequence that complies with every condition, which is performed in algorithm \ref{alg:conditionsmatch}. Hence, it is not necessary to compute $S$ as seen in algorithm \ref{alg:buildsequences}, and with slight modifications we can use the latter to create only one sequence at a time.

\begin{algorithm}[ht]
  \caption{\label{alg:conditionsmatch}\ic{conditionsMatch$(conds, S, ml)$}}
    \KwIn{$conds$ - a vector of conditions\;
      \hspace{1.11cm}$ml$ - an MLabel.}
    \KwOut{the verification result - a boolean.}
    \lIf{$conds=\emptyset$}{\Return \ic{true}\;}
    let $sq=$ \ic{getNextSeq$()$}\;
    let \ic{proceed} $=$ \ic{false}\;
    \For{$i=0$ \KwTo $\left|conds\right|-1$}{
      \While{$sq\neq\emptyset\ \mathbf{and}\ $proceed $=$ false}{
        \ic{proceed} $=$ \ic{false}\;
        \uIf{evaluateCond$(ml,i,sq)=$ false}{
          $sq=$ \ic{getNextSeq$()$}\;
          $i=0$\;
        }
        \lElse{\ic{proceed} $=$ \ic{true}\;}
      }
      \lIf{proceed $=$ false}{\Return \ic{false}\;}
    }
    \Return \ic{true}\;
\end{algorithm}

The outer loop scans the conditions while the inner loop processes the sequences. If a condition evaluation fails (l. 7), the next sequence is requested, and we have to return to the first condition (l. 9) and to request the next sequence (l. 8; this function either returns the next sequence, or, if no more sequences are available, an empty set). Only if an evaluation is successful, the sequence remains a solution candidate, and we proceed to the next condition. Algorithm \ref{alg:conditionsmatch} returns \ic{false} if and only if there is no sequence returned by \ic{getNextSeq$()$} which fulfills all conditions.

Due to lack of space, we limit ourselves to discuss the basic strategies of the evaluation process. In line 7, the function \ic{evaluateCond} receives the moving label, the current sequence, and the number of the current condition. Its major task is to replace parts of the condition according to the sequence. For example, assume the query \texttt{X () Y $\ast$ Z $\ast$ // Y.card < Z.card}, the strongly simplified moving label \texttt{((00-08 at\_home)} \texttt{(08-17 at\_work)} \texttt{(17-19 shopping)} \texttt{(19-00 at\_home))}, and the sequence $\{0,1,4\}$. The latter implies that \texttt{X} corresponds to unit label 0, \texttt{Y} is bound to the other three unit labels while \texttt{Z} remains empty. Now, the expressions \texttt{Y.card} and \texttt{Z.card} from the condition are replaced by \texttt{3} and \texttt{0}, respectively, yielding a negative answer from the \secondo\ evaluation. The next sequence could read $\{0,1,2\}$ -- binding \texttt{Y} to unit label 1 and \texttt{Z} to 2 and 3 -- which leads to the inequation $1<2$, so this sequence is the one fulfilling all conditions.

@@

@@

The filtering works similarly to algorithm \ref{alg:conditionsmatch}, i.e., for every sequence in $S$, we check whether it fulfills every condition. If it does, it stays in $S$, otherwise it is deleted. Now, all information for rewriting the pattern is available.
\newpage

\subsubsection{Display Equations}
A numbered display equation -- one set off by vertical space
from the text and centered horizontally -- is produced
by the \textbf{equation} environment. An unnumbered display
equation is produced by the \textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols
and structures available in \LaTeX; this section will just
give a couple of examples of display equations in context.
First, consider the equation, shown as an inline equation above:
\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
and follow it with another numbered equation:
\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

@@

\subsection{Theorem-like Constructs}
Other common constructs that may occur in your article are
the forms for logical constructs like theorems, axioms,
corollaries and proofs.  There are
two forms, one produced by the
command \texttt{{\char'134}newtheorem} and the
other by the command \texttt{{\char'134}newdef}; perhaps
the clearest and easiest way to distinguish them is
to compare the two in the output of this sample document:

This uses the \textbf{theorem} environment, created by
the\linebreak\texttt{{\char'134}newtheorem} command:
\newtheorem{theorem}{Theorem}
\begin{theorem}
Let $f$ be continuous on $[a,b]$.  If $G$ is
an antiderivative for $f$ on $[a,b]$, then
\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
\end{theorem}

The other uses the \textbf{definition} environment, created
by the \texttt{{\char'134}newdef} command:
\newdef{definition}{Definition}
\begin{definition}
If $z$ is irrational, then by $e^z$ we mean the
unique number which has
logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
\end{definition}

Two lists of constructs that use one of these
forms is given in the
\textit{Author's  Guidelines}.

There is one other similar construct environment, which is
already set up
for you; i.e. you must \textit{not} use
a \texttt{{\char'134}newdef} command to
create it: the \textbf{proof} environment.  Here
is a example of its use:
\begin{proof}
Suppose on the contrary there exists a real number $L$ such that
\begin{displaymath}
\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
\end{displaymath}
Then
\begin{displaymath}
l=\lim_{x\rightarrow c} f(x)
= \lim_{x\rightarrow c}
\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
\frac{f(x)}{g(x)} = 0\cdot L = 0,
\end{displaymath}
which contradicts our assumption that $l\neq 0$.
\end{proof}

Complete rules about using these environments and using the
two different creation commands are in the
\textit{Author's Guide}; please consult it for more
detailed instructions.  If you need to use another construct,
not listed therein, which you want to have the same
formatting as the Theorem
or the Definition\cite{salas:calculus} shown above,
use the \texttt{{\char'134}newtheorem} or the
\texttt{{\char'134}newdef} command,
respectively, to create it.

@@

%ACKNOWLEDGMENTS are optional
%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
\appendix
%Appendix A
\section{Headings in Appendices}
The rules about hierarchical headings discussed above for
the body of the article are different in the appendices.
In the \textbf{appendix} environment, the command
\textbf{section} is used to
indicate the start of each Appendix, with alphabetic order
designation (i.e. the first is A, the second B, etc.) and
a title (if you include one).  So, if you need
hierarchical structure
\textit{within} an Appendix, start with \textbf{subsection} as the
highest level. Here is an outline of the body of this
document in Appendix-appropriate form:
\subsection{Introduction}
\subsection{The Body of the Paper}
\subsubsection{Type Changes and  Special Characters}
\subsubsection{Math Equations}
\paragraph{Inline (In-text) Equations}
\paragraph{Display Equations}
\subsubsection{Citations}
\subsubsection{Tables}
\subsubsection{Figures}
\subsubsection{Theorem-like Constructs}
\subsubsection*{A Caveat for the \TeX\ Expert}
\subsection{Conclusions}
\subsection{Acknowledgments}
\subsection{Additional Authors}
This section is inserted by \LaTeX; you do not insert it.
You just add the names and information in the
\texttt{{\char'134}additionalauthors} command at the start
of the document.
\subsection{References}
Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
%\balancecolumns % GM June 2007
\end{document}

>>>>

\subsubsection{Class NFA}
The NFA instance (there is only one) is constructed in several steps. First, the transition function $\delta$ is computed. We model $\delta$ as an array of mappings from an integer to a set of integers, i.e., for every pair $(i,j)$ of an NFA state and a unit pattern id, there is a (possibly empty) set of states that become active if a unit label matching the unit pattern $j$ is read in state $i$ (replacing the formerly active states). After this, a vector of sets of numbers is filled with all proper matches, e.g., if unit pattern 5 matches unit label 26, the number 26 is inserted into the 5th set.

@@

Concerning the runtime complexity of algorithm \ref{alg:buildNFA}, we observe that it is at least linear in $n$ (since the outer loop processes all $n$ unit patterns) and that it increases with the number of consecutive stars. More exactly, the amortized complexity is $\mathcal O(n)$ if and only if the pattern does not contain sequences of two or more stars. In this case, the CPU cost saved for the star units is roughly spent at the respective subsequent positions. If there are sequences of two or more stars in the pattern, we have to add $\mathcal O(s_i^3)$ for each star sequence, where $s_i$ is the length of the $i$th star sequence. Consequently, for the total complexity of algorithm \ref{alg:buildNFA}, we obtain $\mathcal O(n+\sum_{i=1}^k{s_i^3})$ ($k$ being the number of star sequences).

@@

The CPU time of algorithm \ref{alg:match} depends on the permanently changing numbers of active states and of corresponding transitions. In the worst case, i.e., if every unit pattern is a $\ast$, it amounts to $\mathcal O(mn^2)$ (with $m=\left|ml\right|$). Actually, considerably less than $n^2$ calculations have to be made, because
\begin{itemize}
  \item few states are active simultaneously: $\left|S_{cur}\right|\ll n$,
  \item states have few transitions: $\delta_{s,i}=\emptyset$ for most $i<n$,
  \item the procedure terminates early if $S_{cur}=\emptyset$.
\end{itemize}

The matching decision requested in line 6 is computed in constant time. Hence, if we assume the number of active states and possible transitions to be $\mathcal O(\log n)$, the complexity for algorithm \ref{alg:match} is $\mathcal O(m(\log n)^2)$.

\subsubsection{Preparing the Condition Evaluation}\label{sec:preparing the condition evaluation}
Since the conditions are evaluated by \secondo, there are only two rules they have to obey:
\begin{itemize}
  \item At least one term containing a variable and a condition type (e.g., \texttt{Y.time}) is mandatory for every condition.
  \item Every condition must be a boolean expression, i.e., its evaluation has to yield \ic{true} or \ic{false}.
\end{itemize}

@@

For realizing this, we need to compute a set of cardinality candidates $C_i$ for every unit pattern id $i$, which is done with the help of the matching sets ($M_s$ in algorithm \ref{alg:match}). We distinguish three cases:
\begin{itemize}
  \item a simple pattern: $C_i=\{1\}$,
  \item a sequence pattern surrounded by two proper matches: $C_i=M_{i+1}-M_{i-1}$\footnote{for two sets $A, B\subseteq \mathbb{N}_0$, we denote the set of all non-negative differences $\{b-a\colon a\in A, b\in B, a\leq b\}$ by $B-A$},
  \item a sequence pattern besides other sequence patterns: $C_{i}=\{1,\dots,\max\{M_{i\uparrow}-M_{i\downarrow}\}\}$\footnote{the next and the previous position of simple patterns (starting from $i$) are denoted by $i\!\uparrow$ and $i\!\downarrow$, respectively}.
\end{itemize}

@@

These candidate sets enable us to build further sets providing an evaluation order, as done in algorithm \ref{alg:buildsequences}. First, all cardinality set sizes except the largest one are multiplied, yielding the number of matching sequence candidates which are going to be built in the next step. We skip the largest set in the multiplication since a matching sequence is already determined by $n-1$ numbers. This technique reduces the computational cost for the outer loop by a factor up to $m$, depending on the occurrence of $+$ and $\ast$ in the pattern.

\begin{algorithm}[ht]
  \caption{\label{alg:buildsequences}\ic{buildSequences$(C)$}}
    \KwIn{$C$ - a vector of sets containing the cardinality candidates for every unit pattern.}
    \KwOut{$S$ - a set of multisets representing all possible matching sequences.}
    let $\mu\in \{0,\dots,n-1\}$, with $\left|C_\mu\right|=\max_{0\leq i<n} \left|C_i\right|$\;
    let $size=\prod_{0\leq i<n,i\neq\mu}\left|C_i\right|$\;
    \For{$i=0$ \KwTo $size-1$}{
      let $j=i$\;
      let $sq$ be a multiset of integers, initially empty\;
      let $cards$ be a vector of $n$ integers, initially empty\;
      let $\sigma=0$\;
      \ForEach{$s\in\{0,\dots,n-1\}\setminus \{\mu\}$}{
        $cards_s=C_{s,\,j\ \textnormal{mod}\ \left|C_s\right|}$\;
        $\sigma=\sigma+C_{s,\,j\ \textnormal{mod}\ \left|C_s\right|}$\;
        $j=j/\left|C_s\right|$\;
        \lIf(\tcp*[f]{\scriptsize{sum exceeds limit}}){$\sigma>m$}{$s=n$}
      }
      \If{$m-\sigma\in C_\mu$}{
        $cards_\mu=m-\sigma$\;
        $sq=\{0,\sum_{k=0}^{0}cards_k,\dots,\sum_{k=0}^{n-2}cards_k\}$\;
        $S\leftarrow sq$\;
      }
    }
\end{algorithm}

@@

Due to algorithm \ref{alg:buildsequences}'s potentially huge time consumption -- $\mathcal O(nm^{\omega-1})$, $\omega$ being the number of sequence patterns or 1 if there is none -- this version is only executed if the operator \texttt{rewrite} is called. The other case is discussed subsequently.

\subsubsection{Condition Evaluation}\label{sec:condition evaluation}
The next step for \texttt{rewrite} is to scan $S$, deleting the sequences which do not fulfill the set of conditions in order to obtain the final set of sequences for rewriting the pattern.

However, if the operator \texttt{matches} is invoked, it suffices to verify whether there is a sequence that complies with every condition, which is performed in algorithm \ref{alg:conditionsmatch}. Hence, it is not necessary to compute $S$ as seen above, and we can modify algorithm \ref{alg:buildsequences} to a function \ic{getNextSeq$()$} returning one sequence in $\mathcal O(n)$.

\begin{algorithm}[ht]
  \caption{\label{alg:conditionsmatch}\ic{conditionsMatch$(conds, S, ml)$}}
    \KwIn{$conds$ - a vector of $c$ conditions\;
      \hspace{1.11cm}$ml$ - an MLabel.}
    \KwOut{the verification result - a boolean.}
    \lIf{$conds=\emptyset$}{\Return \ic{true}\;}
    let $sq=$ \ic{getNextSeq$()$}\;
    let \ic{proceed} $=$ \ic{false}\;
    \For{$i=0$ \KwTo $c-1$}{
      \While{$sq\neq\emptyset\ \mathbf{and}\ $proceed $=$ false}{
        \uIf{evaluateCond$(ml,i,sq)=$ false}{
          $sq=$ \ic{getNextSeq$()$}\;
          $i=0$\;
        }
        \lElse{\ic{proceed} $=$ \ic{true}\;}
      }
      \lIf{proceed $=$ false}{\Return \ic{false}\;}
    }
    \Return \ic{true}\;
\end{algorithm}

The outer loop scans the conditions while the inner loop processes the sequences. If a condition evaluation fails (l. 6), the next sequence is requested, and we have to return to the first condition (l. 8) and to request the next sequence (l. 7; this function either returns the next sequence, or, if no more sequences are available, an empty set). Only if an evaluation is successful, the sequence remains a solution candidate, and we proceed to the next condition. Algorithm \ref{alg:conditionsmatch} returns \ic{false} if and only if there is no sequence returned by \ic{getNextSeq$()$} which fulfills all conditions.

Since \secondo\ evaluation details are beyond the scope of this paper, we discuss merely the basic concept of the evaluation process. In line 6, the function \ic{evaluateCond} receives the moving label, the current sequence, and the number of the current condition. Its major task is to replace parts of the condition according to the sequence. For example, assume the query \texttt{X () Y $\ast$ Z $\ast$ // Y.card < Z.card}, the strongly simplified moving label \texttt{((00-08 at\_home)} \texttt{(08-17 at\_work)} \texttt{(17-19 shopping)} \texttt{(19-00 at\_home))}, and the sequence $\{0,1,4\}$. The latter implies that \texttt{X} corresponds to unit label 0, \texttt{Y} is bound to the other three unit labels while \texttt{Z} remains empty. Now, the expressions \texttt{Y.card} and \texttt{Z.card} from the condition are replaced by \texttt{3} and \texttt{0}, respectively, yielding a negative answer from the \secondo\ evaluation. The next sequence could read $\{0,1,2\}$ -- binding \texttt{Y} to unit label 1 and \texttt{Z} to 2 and 3 -- which leads to the inequation $1<2$, so this sequence is the one fulfilling all conditions.

@@

The worst case complexity for algorithm \ref{alg:conditionsmatch} is $\mathcal O(cnm^{\omega-1})$ ($\omega$ as in \ref{sec:preparing the condition evaluation}) Again, the actual runtime is far below since
\begin{itemize}
  \item only if condition $i-1$ is fulfilled, the $i$th one is tested; assuming half of the conditions to be true results in an expectation of $\sum_{i=0}^{c-1}1/2^i$ conditions, i.e., at most 2,
  \item not all sequences must always be computed.
\end{itemize}

For the average case, we obtain a $\mathcal O(nm^{\omega-1})$ complexity.

@@

The filtering works similarly to algorithm \ref{alg:conditionsmatch}, i.e., for every sequence in $S$, we check whether it fulfills every condition. If it does, it stays in $S$, otherwise it is deleted. Now, all information for rewriting the pattern is available. From each remaining sequence in $S$, we construct a new moving label, depending on the commands given by the user in the result and assignment sections, respectively. This concludes the \texttt{rewrite} operation.

\subsubsection{Linear Runtime}
As seen, several properties of the input have an impact on the runtime. In order to achieve the optimal runtime of $\mathcal O(n+m)$, the only restriction is not to use sequence patterns. For this case, we sum up the time consumption for every step: Input parsing and NFA construction are in $\mathcal O(n)$, the match procedure is in $\mathcal O(m)$, the conditions are evaluated in $\mathcal O(n)$, and the rewriting of the moving label (if \texttt{rewrite} is invoked) is done in $\mathcal O(m)$ since there is only one possible result. Altogether, we have the proposed runtime.

Already the use of one sequence pattern causes a higher complexity, more exactly, the number of states that are active at the same time in algorithm \ref{alg:match} is not constant in this case.

\subsection{Experimental Evaluation}
This section is devoted to the experiments carried out with the two operators. We vary the size of the moving label, the size of the pattern, the number of sequence patterns, and the number of conditions.
\\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ \\ 
\newpage

@@

@@

%ACKNOWLEDGMENTS are optional
%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
%\balancecolumns % GM June 2007
\end{document}
